---
title: "STA2201-Lab-8"
author: "Alice Huang"
date: "22/03/2023"
output:
  pdf_document: default
  html_document: default
---

# Child mortality in Sri Lanka

In this lab you will be fitting a couple of different models to the data about child mortality in Sri Lanka, which was used in the lecture. Here's the data and the plot from the lecture:

```{r, include = FALSE}
library(tidyverse)
library(here)
library(rstan)
library(tidybayes)
```
```{r}
lka <- read_csv(here("lka.csv"))
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```

# Fitting a linear model 

Let's firstly fit a linear model in time to these data. Here's the code to do this:

```{r}
observed_years <- lka$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = mean(years), se = lka$se)

mod <- stan(data = stan_data, file = here("lka_linear_me.stan"))
```

Extract the results:

```{r}
res <- mod %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year = years[t])
```


Plot the results:

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res, aes(year, .value)) + 
  geom_ribbon(data = res, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Linear fit shown in black")
```

## Question 1

Project the linear model above out to 2023 by adding a `generated quantities` block in Stan (do the projections based on the expected value $\mu$). Plot the resulting projections on a graph similar to that above. 

```{r}
stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = mean(years), se = lka$se, P=9)

mod2 <- stan(data = stan_data,
             file = here("lab10.stan"))

```


```{r}
res2 <- mod2 %>%
   gather_draws(mu[t]) %>%
  median_qi() %>%
  mutate(year = years[t])

res_p2 <- mod2 %>%
  gather_draws(mu_p[p]) %>%
  median_qi() %>%
  mutate(year = years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  geom_line(data=res2, aes(year, .value)) +
  geom_ribbon(data=res2, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_line(data = res_p2, aes(year, .value), col = "red") +
  geom_ribbon(data=res_p2, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "red") +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio", subtitle = "Linear fit shown in black")
```

## Question 2

Code up and estimate a first order random walk model to fit to the Sri Lankan data, taking into account measurement error, and project out to 2023. 

```{r}
stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  se = lka$se, P=9)


mod3 <- stan(data = stan_data,
             file = here("lab10_2.stan"))

```

```{r}
res3 <- mod3 %>%
   gather_draws(mu[t]) %>%
  median_qi() %>%
  mutate(year = years[t])

res_p3 <- mod3 %>%
  gather_draws(mu_p[p]) %>%
  median_qi() %>%
  mutate(year = years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  geom_line(data=res3, aes(year, .value)) +
  geom_ribbon(data=res3, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_line(data = res_p3, aes(year, .value), col = "red") +
  geom_ribbon(data=res_p3, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "red") +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```


## Question 3

Now alter your model above to estimate and project a second-order random walk model (RW2). 

```{r}
stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  se = lka$se, P=9)


mod4 <- stan(data = stan_data,
             file = here("lab10_3.stan"))

```

```{r}
res4 <- mod4 %>%
   gather_draws(mu[t]) %>%
  median_qi() %>%
  mutate(year = years[t])

res_p4 <- mod4 %>%
  gather_draws(mu_p[p]) %>%
  median_qi() %>%
  mutate(year = years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  geom_line(data=res4, aes(year, .value)) +
  geom_ribbon(data=res4, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_line(data = res_p4, aes(year, .value), col = "red") +
  geom_ribbon(data=res_p4, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "red") +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```


## Question 4

Run the first order and second order random walk models, including projections out to 2023. Compare these estimates with the linear fit by plotting everything on the same graph. 

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  geom_line(data=res2, aes(year, .value), col="blue") +
  geom_ribbon(data=res2, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, colour = "blue") +
  geom_line(data = res_p2, aes(year, .value), col = "blue") +
  geom_ribbon(data=res_p2, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "blue") +
  geom_line(data=res3, aes(year, .value), col="green") +
  geom_ribbon(data=res3, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill="green") +
  geom_line(data = res_p3, aes(year, .value), col = "green") +
  geom_ribbon(data=res_p3, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "green") +
  geom_line(data=res4, aes(year, .value), col="red") +
  geom_ribbon(data=res4, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, col="red") +
  geom_line(data = res_p4, aes(year, .value), col = "red") +
  geom_ribbon(data=res_p4, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "red") +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio", subtitle = "Red = 2nd order random walk, Green = 1st order random walk, Blue = linear")
```


## Question 5

Rerun the RW2 model excluding the VR data. Briefly comment on the differences between the two data situations. 

First we filter out the VR dataset.

```{r}
lkavr <- lka %>% filter(source != "VR")
observed_yearsvr <- lkavr$year
yearsvr <- min(observed_yearsvr):max(observed_yearsvr)
nyearsvr <- length(yearsvr)

stan_datavr <- list(y = lkavr$logit_ratio, year_i = observed_yearsvr - yearsvr[1]+1, 
                  T = nyearsvr, years = yearsvr, N = length(observed_yearsvr), 
                  se = lkavr$se, P=9)


mod4 <- stan(data = stan_datavr,
             file = here("lab10_3.stan"))

```


```{r}
res4 <- mod4 %>%
   gather_draws(mu[t]) %>%
  median_qi() %>%
  mutate(year = yearsvr[t])

res_p4 <- mod4 %>%
  gather_draws(mu_p[p]) %>%
  median_qi() %>%
  mutate(year = yearsvr[nyearsvr]+p)
```

```{r}
ggplot(lkavr, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  geom_line(data=res4, aes(year, .value)) +
  geom_ribbon(data=res4, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_line(data = res_p4, aes(year, .value), col = "red") +
  geom_ribbon(data=res_p4, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "red") +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```

The VR source is the only one that has the most recent observation in 2014. If we exclude VR, the next most recent observation is from 2005. Since the second order random walk projection depends on the most recent observation and second most recent observation, excluding VR will make the projections be based off older observations, and potentially be less accurate. We see that the projected 2023 value is lower if we exclude the VR source.

## Question 6

Briefly comment on which model you think is most appropriate, or an alternative model that would be more appropriate in this context. 

We run the random walk order 1 model on the data excluding VR.

```{r}
mod5 <- stan(data = stan_datavr,
             file = here("lab10_2.stan"))
```

```{r}
res5 <- mod5 %>%
   gather_draws(mu[t]) %>%
  median_qi() %>%
  mutate(year = yearsvr[t])

res_p5 <- mod5 %>%
  gather_draws(mu_p[p]) %>%
  median_qi() %>%
  mutate(year = yearsvr[nyearsvr]+p)
```

```{r}
ggplot(lkavr, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  geom_line(data=res4, aes(year, .value)) +
  geom_ribbon(data=res4, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_line(data = res_p4, aes(year, .value), col = "red") +
  geom_ribbon(data=res_p4, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "red") +
  geom_line(data=res5, aes(year, .value), col="blue") +
  geom_ribbon(data=res5, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "green") +
  geom_line(data = res_p5, aes(year, .value), col = "blue") +
  geom_ribbon(data=res_p5, aes(y= .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = "green") +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```

I think the second order random walk model is more appropriate for data excluding the VR source. Since the first order random walk projection depends on the most recent observation, excluding VR will make the projections be based off only one observation from 2005. The first order random walk projection assumes the projected values will be similar to the 2005 values. However the second order random walk projection will depend on the observations from 2005 and 2000 so perhaps it captures the trend better. The second order random walk model gives more reasonable point estimates